# Data preparation
This notebook prepares the publicly available RNA-seq count data from GSE260786 for downstream analysis.

# Load data
```{r}
library(readxl)
library(dplyr)
library(tibble)
library(here)

raw_data <- read_excel(here("data", "GSE260786_bulk.xlsx"))

dim(raw_data)
colnames(raw_data)[1:10]
```

# Prepare sample metadata
```{r}
metadata <- data.frame(Sample = colnames(raw_data)[-1], Group = as.character(raw_data[1, -1]),stringsAsFactors = FALSE)

table(metadata$Group)

# Subset only IIM and Control
metadata_iim <- metadata %>% filter(Group %in% c("ASYS", "DM", "IBM", "IMNM", "Control")) %>%
  mutate(analysis_group = ifelse(Group == "Control", "Control", "IIM"))
```

# Prepare count matrix
Rows without valid Ensembl gene identifiers were excluded. Ensembl gene IDs were then used as row names, read counts were converted to numeric values, and genes containing missing count values were removed.
```{r}
raw_counts <- raw_data[-1, ] %>% filter(grepl("^ENSG", Geneid)) %>% column_to_rownames("Geneid") %>%mutate(across(everything(), as.numeric))

#remove genes with missing count values
raw_counts <- raw_counts[rowSums(is.na(raw_counts)) == 0,]

#subset count matrix to IIM and Control samples
raw_counts_iim <- raw_counts[, metadata_iim$Sample]

#check sample order
identical(colnames(raw_counts_iim), metadata_iim$Sample) #TRUE
```

# Validate count matrix
Basic checks were performed to ensure that the count matrix contained unique gene identifiers and valid non-negative integer counts before downstream analysis.
```{r}
#check duplicated gene IDs
sum(duplicated(rownames(raw_counts_iim))) #0

#check missing values
sum(is.na(raw_counts_iim)) #0

#check negative counts
sum(as.matrix(raw_counts_iim) < 0) #0

#check non-integer counts
sum(as.matrix(raw_counts_iim) %% 1 != 0) #0
```

# Save
```{r}
saveRDS(raw_counts_iim,here("data", "raw_counts_iim_control.rds"))

saveRDS(metadata_iim,here("data", "metadata_iim_control.rds"))
```


The cleaned count matrix and sample metadata were successfully prepared for downstream analysis.
